---
title: "Elementary Probability Theory"
subtitle: "Solved Problems"
author: "Joseph Bennett"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Fundamental rules of probability

__6.1.__ Determine the probability $p$, or an estimate of it, for each of the
following events:

(a) An odd number appears in a single toss of a fair die.
(b) At least one head appears in two tosses of a fair coin.
(c) An ace, 10 of diamonds, or 2 of spades appears in drawing a single card from
a well-shuffled ordinary deck of 52 cards.
(d) The sum 7 appears in a single toss of a pair of fair dice.
(e) A tail appears in the next toss of a coin if out of 100 tosses 56 were
heads.

__Solution__

(a) Out of six possible equally likely cases, three cases (where the die comes
up 1, 3, or 5) are favorable to the event. Thus $p = \frac{3}{6} = \frac{1}{2}$.
(b) If $H$ denotes "head" and $T$ denotes "tail," the two tosses can lead to
four cases: $HH$, $HT$, $TH$, and $TT$, all equally likely. Only the first three
cases are favorable to the event. Thus $p = \frac{3}{4}$.
(c) The event can occur in six ways (ace of spades, ace of hearts, ace of clubs,
ace of diamonds, 10 of diamonds, and 2 of spades) out of 52 equally likely
cases. Thus $p = \frac{6}{52} = \frac{3}{26}$.
(d) Each of the six faces of one die can be associated with each of the six
faces of the other die, so that the total number of cases that can arise, all
equally likely, is $6 \cdot 6 = 36$. These can be denoted by (1,1), (2,1),
(3,1), \ldots, (6,6).
(e) Since 100 - 56 = 44 tails were obtained in 100 tosses, the _estimated_ 
(or _empirical_) _probability_ of a tail is the relative frequency
44/100 = 0.44.

---

__6.2.__ An experiment consists of tossing a coin and a die. If $E_1$ is the
event that "head" comes up in tossing the coin and $E_2$ is the event that
"3 or 6" comes up in tossing the die, state in words the meaning of each of
the following:

(a) $\bar{E_1}$
(b) $\bar{E_2}$
(c) $E_1 E_2$
(d) $\text{Pr}\{E_1 E_2\}$
(e) $\text{Pr}\{E_1|E_2\}$
(f) $\text{Pr}\{E_1 + E_2\}$

__Solution__

(a) Tails on the coin and anything on the die
(b) 1, 2, 4, or 5 on the die and anything on the coin
(c) Heads on the coin and 3 or 6 on the die
(d) Probability of heads on the coin and 1, 2, 4, or 5 on the die
(e) Probability of heads on the coin, given that a 3 or 6 has come up on the die
(f) Probability of tails on the coin or 1, 2, 4, or 5 on the die, or both

---

__6.3.__ A ball is drawn at random from a box containing 6 red balls, 4 white
balls, and 5 blue balls. Determine the probability that the ball drawn is (a)
red, (b) white, (c) blue, (d) not red, and (e) red or white.

__Solution__

Let $R$, $W$, and $B$ denote the events of drawing a red ball, a white ball, and
blue ball, respectively. Then:

(a)
$$
\text{Pr}\{R\} = \frac{\text{ways of choosing a red ball}}
                      {\text{total ways of choosing a ball}}
  = \frac{6}{6 + 4 + 5} = \frac{6}{15} = \frac{2}{5}
$$

(b)
$$
\text{Pr}\{W\} = \frac{4}{6 + 4 + 5} = \frac{4}{15}
$$

(c)
$$
\text{Pr}\{B\} = \frac{5}{6 + 4 + 5} = \frac{5}{15} = \frac{1}{3}
$$

(d)
$$
\text{Pr}\{\bar{B}\} = 1 - \text{Pr}\{R\} = 1 - \frac{2}{5} = \frac{3}{5}
\ \ \ \text{by part (a)}
$$

(e)
$$
\text{Pr}\{R + W\} = \frac{\text{ways of choosing a red or white ball}}
                          {\text{total ways of choosing a ball}}
  = \frac{6 + 4}{6 + 4 + 5} = \frac{10}{15} = \frac{2}{3}
$$

__Another method__

$$
\text{Pr}\{R + W\} = \text{Pr}\{\bar{B}\} = 1 - \text{Pr}\{B\}
  = 1 - \frac{1}{3} = \frac{2}{3}\ \ \ \text{by part (c)}
$$

Note that $\text{Pr}\{R + W\} = \text{Pr}\{R\} + \text{Pr}\{W\}$ (i.e.,
$\frac{2}{3} = \frac{2}{5} + \frac{4}{15}$). This is an illustration of the
general rule $\text{Pr}\{E_1 + E_2\} = \text{Pr}\{E_1\} + \text{Pr}\{E_2\}$
that is true for _mutually exclusive_ events $E_1$ and $E_2$.

---

__6.4.__ A fair die is tossed twice. Find the probability of getting a 4, 5, or
6 on the first toss and a 1, 2, 3 or 4 on the second toss.

__Solution__

Let $E_1 =$ event "4, 5, or 6" on the first toss and let $E_2 =$ event "1, 2, 3,
or 4" on the second toss. Each of the six ways 
in which the die can fall on the first toss can be associated with each of the
six ways in which it can fall on the second toss, a total of $6 \cdot 6 = 36$
ways, all equally likely. Each of the three ways in which $E_1$ can occur can
be associated with each of the four ways in which $E_2$ can occur, to give
$3 \cdot 4 = 12$ ways in which both $E_1$ and $E_2$, or $E_1 E_2$ occur. Thus
$\text{Pr}\{E_1 E_2\} = 12/36 = 1/3$.

Note that $\text{Pr}\{E_1 E_2\} = \text{Pr}\{E_1\}\ \text{Pr}\{E_2\}$ (i.e.,
$\frac{1}{3} = \frac{3}{6} \cdot \frac{4}{6}$) is valid for the
_independent events_ $E_1$ and $E_2$.

---

__6.5.__ Two cards are drawn from a well-shuffled ordinary deck of 52 cards.
Find the probability that they are both aces if the first card is (a) replaced
and (b) not replaced.

__Solution__

Let $E_1 =$ event "ace" on the first draw, and $E_2 =$ event "ace" on the second
draw.

(a) If the first card is replaced, then $E_1$ and $E_2$ are independent events.
Thus Pr{both cards are aces} = Pr{$E_1 E_2$} = Pr{$E_1$} Pr{$E_2$} =
$\left(\frac{4}{52}\right) \left(\frac{4}{52}\right) = \frac{1}{169}$.
(b) The first card can be drawn in any one of 52 ways, and the second card can
be drawn in any one of 51 ways since the first card is not replaced. Thus both
cards can be drawn in $52 \cdot 51$ ways, all equally likely.

> There are four ways in which $E_1$ can occur and three ways in which $E_2$ can
occur, so that both $E_1$ and $E_2$, or $E_1 E_2$, can occur in $4 \cdot 3$
ways. Thus Pr{$E_1 E_2$} =
$\left(4 \cdot 3 \right)/ \left(52 \cdot 51 \right) = \frac{1}{221}$.

> Note that Pr{$E_2|E_1$} = Pr{second card is an ace given that first card is an
ace} = $\frac{3}{51}$. Thus our result is an illustration of the general rule
that Pr{$E_1 E_2$} = Pr{$E_1$} Pr{$E_2|E_1$} when $E_1$ and $E_2$ are dependent
events.

---

__6.6.__ Three balls are drawn successively from the box of Problem 6.3. Find
the probability that they are drwan in the order red, white, and blue if each
ball is (a) replaced and (b) not replaced.

__Solution__

Let $R =$ event "red" on the first draw, $W =$ event "white" on the second draw,
and $B =$ event "blue" on the third draw. We require Pr{$RWB$}.

(a) If each ball is replaced, then $R$, $W$, and $B$ are independent events and

$$
\text{Pr \{RWB\} = Pr\{R\} Pr\{W\} Pr\{B\}}
  = \left( \frac{6}{6 + 4 + 5} \right) \left( \frac{4}{6 + 4 + 5} \right)
    \left( \frac{5}{6 + 4 + 5} \right)
  = \left(\frac{6}{15}\right) \left(\frac{4}{15}\right)
    \left(\frac{5}{15}\right) = \frac{8}{225}
$$

(b) If each ball is not replaced, then R, W, and B are dependent events and

$$
\text{Pr\{RWB\} = Pr\{R\} Pr\{W|R\} Pr\{B|WR\}}
  = \left(\frac{6}{6 + 4 + 5}\right) \left(\frac{4}{5 + 4 + 5}\right)
    \left(\frac{5}{5 + 3 + 5}\right)
  = \left(\frac{6}{15}\right) \left(\frac{4}{14}\right)
    \left(\frac{5}{13}\right) = \frac{4}{91}
$$

> where Pr{B|WR} is the conditional probability of getting a blue ball if a
white and red ball have already been chosen.

---

__6.7.__ Find the probabilty of a 4 turning up at least once in two tosses of a
fair die.

__Solution__

Let $E_1$ = event "4" on the first toss, $E_2$ = event "4" on the second toss,
and $E_1 + E_2$ = event "4" on the first toss or "4" on the second toss or both
= event that at least one 4 turns up. We require Pr{$E_1 + E_2$}.

__First method__

The total number or equally likely ways in which both dice can fall is
$6 \cdot 6 = 36$. Also,

$$
\text{Number of ways in which } E_1 \text{ occurs but not } E_2 = 5 \\
\text{Number of ways in which } E_2 \text{ occurs but not } E_1 = 5 \\
\text{Number of ways in which both } E_1 \text{ and } E_2 \text{ occur } = 1
$$

Thus the number of ways in which at least one of the events $E_1$ or $E_2$
occurs is $5 + 5 + 1 = 11$, and thus Pr{$E_1 + E_2$} = $\frac{11}{36}$.

__Second method__

Since $E_1$ and $E_2$ are not mutually exclusive, Pr{$E_1 + E_2$} = Pr{$E_1$}
+ Pr{$E_2$} - Pr{$E_1 E_2$}. Also, since $E_1$ and $E_2$ are independent,
Pr{$E_1 E_2$} = Pr{$E_1$} Pr{$E_2$}. Thus Pr{$E_1 + E_2$} = Pr{$E_1$}
+ Pr{$E_2$} - Pr{$E_1$} Pr{$E_2$} =
$\frac{1}{6} + \frac{1}{6} = \left(\frac{1}{6}\right) \left(\frac{1}{6}\right)$
$= \frac{11}{36}$.

__Third method__

$$
\text{Pr\{at least one 4 comes up\}} + \text{Pr\{no 4 comes up\}} = 1
$$

Thus
$$
\begin{aligned}
\text{Pr\{at least one 4 comes up\}} & = 1 - \text{Pr\{no 4 comes up\}} \\
  & = 1 - \text{Pr\{no 4 on first toss and no 4 on second toss\}} \\
  & = 1 - \text{Pr}\{\bar{E_1} \bar{E_2}\}
       = 1 - \text{Pr}\{\bar{E_1}\}\ \text{Pr}\{\bar{E_2}\} \\
  & = 1 - \left(\frac{5}{6}\right) \left(\frac{5}{6}\right) = \frac{11}{36}
\end{aligned}
$$

---

__6.8.__ One bag contains 4 white balls and 2 black balls; another contains 3
white balls and 5 black balls. If one ball is drawn from each bag, find the
probability that (a) both are white, (b) both are black, and (c) one is white
and one is black.

__Solution__

Let $W_1 =$ event "white" ball from the first bag, and let $W_2 =$ event "white"
ball from the second bag.

(a)
$$
\text{Pr}\{W_1 W_2\} = \text{Pr}\{W_1\}\ \text{Pr}\{W_2\}
 = \left( \frac{4}{4 + 2} \right) \left(\frac{3}{3 + 5}\right) = \frac{1}{4}
$$
 
(b)
$$
\text{Pr}\{\bar{W_1} \bar{W_2}\} = \text{Pr}\{\bar{W_1}\}\ \text{Pr}\{\bar{W_2}\}
  = \left( \frac{2}{4 + 2} \right) \left( \frac{5}{3 + 5} \right) = \frac{5}{24}
$$

(c) The event "one is white and one is black" is the same as the event "either
the first is white and the second is black _or_ the first is black and the
second is white"; that is, $W_1 \bar{W_2} + \bar{W_1} W_2$. Since events
$W_1 \bar{W_2}$ and $\bar{W_1} W_2$ are mutually exclusive, we have

$$
\begin{aligned}
\text{Pr} \{ W_1 \bar{W_2} + \bar{W_1} W_2 \} & = \text{Pr} \{W_1 \bar{W_2} \} 
   + \text{Pr} \{ \bar{W_1} W_2 \} \\
 & = \text{Pr} \{ W_1 \}\ \text{Pr} \{ \bar{W_2} \}
   + \text{Pr} \{ \bar{W_1} \}\ \text{Pr} \{ \bar{W_1} \}\ \text{Pr} \{ W_2 \} \\
 & = \left( \frac{4}{4 + 2} \right) \left( \frac{5}{3 + 5} \right)
   + \left( \frac{2}{4 + 2} \right) \left( \frac{3}{3 + 5} \right)
 = \frac{13}{24}
\end{aligned}
$$

__Another method__

The required probability is
$1 - \text{Pr} \{ W_1 W_2 \} - \text{Pr} \{ \bar{W_1} \bar{W_2} \} =$
$1 - \frac{1}{4} - \frac{5}{24} = \frac{13}{24}$.

---

__6.9.__ $A$ and $B$ play 12 games of chess, of which 6 are won by $A$, 4 are
won by $B$, and 2 end in a draw. They agree to play a match consisting of 3
games. Find the probability that (a) $A$ wins all 3 games, (b) 2 games end in a
draw, (c) $A$ and $B$ win alternately, and (d) $B$ wins at least 1 game.

__Solution__

Let $A_1$, $A_2$, and $A_3$ denote the events "A wins" in the first, second, and
third games, respectively; let $B_1$, $B_2$, and $B_3$ denote the events "B
wins" in the first, second, and third games, respectively; and let, $D_1$,
$D_2$, and $D_3$ denote the events "there is a draw" in the first, second, and
third games, respectively.

On the basis of their past experience (empirical probability), we shall assume
that Pr{$A$ wins any one game} = $\frac{6}{12}$ = $\frac{1}{2}$, that
Pr{$B$ wins any one game} = $\frac{4}{12} = \frac{1}{3}$, and that Pr{any one
game ends in a draw} = $\frac{2}{12} = \frac{1}{6}$.

(a) Pr{$A$ wins all 3 games} = Pr{$A_1 A_2 A_3$} = Pr{$A_1$} Pr{$A_2$} Pr{$A_3$}
= $\left( \frac{1}{2} \right) \left( \frac{1}{2} \right)$
$\left( \frac{1}{2} \right) = \frac{1}{8}$ assuming that the results of each
game are independent of the results of any others, which appears to be
justifiable (unless, of course, the players happen to be
_psychologically influenced_ by the other one's winning or losing).

(b)
$$
\begin{aligned}
\text{Pr\{2 games end in a draw\}}
  & = \text{Pr\{1st and 2nd or 1st and 3rd or 2nd and 3rd games end in a draw\}} \\
  & = \text{Pr}\{D_1 D_2 \bar{D_3}\} + \text{Pr}\{D_1 \bar{D_2} D_3\}
    + \text{Pr}\{\bar{D_1} D_2 D_3\} \\
  & = \text{Pr}\{D_1\}\ \text{Pr}\{D_2\}\ \text{Pr}\{\bar{D_3}\}
    + \text{Pr}\{D_1\}\ \text{Pr}\{\bar{D_2}\}\ \text{Pr}\{D_3\} \\
  & + \text{Pr}\{\bar{D_1}\}\ \text{Pr}\{D_2\}\ \text{Pr}\{D_3\} \\
  & = \left(\frac{1}{6}\right) \left(\frac{1}{6}\right) \left(\frac{5}{6}\right)
    + \left(\frac{1}{6}\right) \left(\frac{5}{6}\right) \left(\frac{1}{6}\right)
    = \frac{15}{216} = \frac{5}{72}
\end{aligned}
$$

(c)
$$
\begin{aligned}
\text{Pr\{A and B win alternately\}}
  & = \text{Pr}\{\text{A wins then B wins then A wins or B wins then A wins then B wins}\} \\
  & = \text{Pr}\{A_1 B_2 A_3 + B_1 A_2 B_3\} = \text{Pr}\{A_1 B_2 A_3\} + \text{Pr}\{B_1 A_2 B_3\} \\
  & = \text{Pr}\{A_1\}\ \text{Pr}\{B_2\}\ \text{Pr}\{A_3\}
    + \text{Pr}\{B_1\}\ \text{Pr}\{A_2\}\ \text{Pr}\{B_3\} \\
  & = \left(\frac{1}{2}\right) \left(\frac{1}{3}\right) \left(\frac{1}{2}\right)
    + \left(\frac{1}{3}\right) \left(\frac{1}{2}\right) \left(\frac{1}{3}\right)
    = \frac{5}{36} \\
\end{aligned}
$$

(d)
$$
\begin{aligned}
\text{Pr\{B wins \t least 1 game\}}
  & = 1 - \text{Pr}\{\text{B wins no game}\} \\
  & = 1 - \text{Pr}\{\bar{B_1} \bar{B_2} \bar{B_3}\} = 1
    - \text{Pr}\{\bar{B_1}\}\ \text{Pr}\{\bar{B_2}\}\ \text{Pr}\{\bar{B_3}\} \\
  & = 1 - \left(\frac{2}{3}\right) \left(\frac{2}{3}\right)
          \left(\frac{2}{3}\right) = \frac{19}{27}
\end{aligned}
$$

---

# Probability distributions

__6.10__ Find the probability of boys and girls in families with three children,
assuming equal probabilities of boys and girls.

__Solution__

Let $B =$ event "boy in the famil," and let $G =$ event "girl in the family."
Thus according to the assumption of equal probabilities, Pr{$B$} = Pr{$G$} = 
$\frac{1}{2}$. In families of three children the following mutually exclusive
events can occur with the corresponding indicated probabilities:

(a) Three boys (BBB):

$$
\text{Pr\{BBB\}} = \text{Pr\{B\} Pr\{B\} Pr\{B\}} = \frac{1}{8}
$$

Here we assume that the birth of a boy is not influenced in any manner by the
fact that a previous child was also a boy, that is, we assume that the events
are independent.

(b) Three girls (GGG): As in part (a) or by symmetry,

$$
\text{Pr\{GGG\}} = \frac{1}{8}
$$

(c) Two boys and one girl (BBG + BGB + GBB):

$$
\begin{aligned}
\text{Pr\{BBG + BGB + GBB\}} & = \text{Pr\{BBG\}} + \text{Pr\{BGB\}}
  + \text{Pr\{GBB\}} \\
  & = \text{Pr\{B\} Pr\{B\} Pr\{G\}} + \text{Pr\{B\} Pr\{G\} Pr\{B\}}
    + \text{Pr\{G\} Pr\{B\} Pr\{B\}} \\
  & = \frac{1}{8} + \frac{1}{8} + \frac{1}{8} = \frac{3}{8}
\end{aligned}
$$

(d) Two girls and one boy (GGB + GBG + BGG): As in part (c) or by symmetry, the
probability is $\frac{3}{8}$.

If we call $X$ the _random variable_ showing the number of boys in families with
three children, the probability distribution is as shown in Table 6.2.

```{r echo=F}
X <- c(0, 1, 2, 3)
p.X <- c("1/8", "3/8", "3/8", "1/8")
table_6.2 <- matrix(c(X,p.X), nrow=2, byrow=T)
rownames(table_6.2) <- c("Number of boys X", "Probability p(X)")
knitr::kable(table_6.2,
             caption = "probability distribution for the number of boys")
```

---

__6.11__ Graph the distribution of Problem 6.10

__Solution__

The graph can be represented either as in Fig. 6-7 or Fig. 6-8. Note that the
sum of the areas of the rectangles in Fig. 6-8 is 1; in this figure, called a
_probability histogram_, we are considering $X$ as a continuous variable even
though it is actually discrete, a procedure that is often found useful.
Figure 6-7, on the other hand, is used when one does not wish to consider the
variable as continuous.

```{r echo=F, fig.cap="Probability distribution"}
plot(c(1/8, 3/8, 3/8, 1/8), xlab="Boys", ylab="Probability", pch=19)
```

```{r echo=F, message=F, fig.cap="probability histogram."}
require(ggplot2)
Percent <- c(1/8, 3/8, 3/8, 1/8)
Boys <- c(0, 1, 2, 3)
df <- data.frame(Boys,Percent)
ggplot(df, aes(x=Boys, y=Percent)) + geom_col()
```

---

__6.12__ A continuous random variable $X$, having values only between 0 and 5,
has a density function given by

$$
p(X) = \begin{cases}
  0.2, 0 < X < 5 \\
    0, \text{otherwise}
\end{cases} \text{.}
$$

The graph is shown in Fig. 6-9.

(a) Verify that it is a density function.

```{r echo=F}
require(ggplot2)
x <- c(0, 5, 5)
y <- c(0.2, 0.2, 0)
df <- data.frame(x,y)
ggplot(df, aes(x=x, y=y, group=1)) + geom_line()
```

(b) Find and graph Pr{2.5 < X < 4.0}.

__Solution__

(a) The function $p(X)$ is always $\ge 0$ and the total area under graph of
$p(X)$ is $5 \times 0.2 = 1$ since it is rectangular in shape and has width 0.2
and length 5 (see Fig. 6-9).

(b) The probability Pr{2.5 < X < 5.0} is shown in Fig. 6-10.


```{r echo=F, fig.cap="The probability Pr{2.5 <  X 4.0} is shown a the blue area."}
require(ggplot2)
x <- c(0, 0,   5,   5)
y <- c(0, 0.2, 0.2, 0)
df <- data.frame(x,y)
ggplot(df, aes(x=x, y=y, group=1)) +
  geom_line() +
  annotate('rect', colour='black', fill='lightblue',
           xmin=2.5, xmax=4, ymin=0, ymax=0.2) +
  ylim(0, 0.250)
```

# Mathematical expectation

__6.13__ If a man purchases a raffle ticket, he can win a first prize of
\$5000 or a second prize of \$2000 with probabilities 0.001 and 0.003. What
should be a fair price to pay for the ticket?

__Solution__

His expectation is $(\$5000)(0.001) + (\$2000)(0.003) = \$5 + \$6 = \$11$,
which is a fair price to pay.

---

__6.14__ In a given business venture a lady can make a profit of \$300 with 
probability 0.6 or take a loss of \$100 with probability 0.4. Determine her
expectation.

__Solution__

Her expectation is $(\$300)(0.6) + (\$100)(0.4) = \$180 - \$40 = \$140$.

---

__6.15__ Find (a) $E(X)$, (b) $E(X^2)$, and (c) $E[(X - \bar{X})^2]$ for the
probability distribution shown in Table 6.3. (d) Give the `R` solution.

```{r}
X <- c(8, 12, 16, 20, 24)
p.X <- c(1/8, 1/6, 3/8, 1/4, 1/12)
table_6.3 <- matrix(c(X, p.X), nrow=2, byrow=T)
rownames(table_6.3) <- c("X", "p(X)")
knitr::kable(table_6.3)
```

__Solution__

We will combine all three:

```{r}
output <- t(rbind(table_6.3, X*p.X, X^2*p.X, (X - sum(X*p.X))^2*p.X))
colnames(output) <- c("X", "p(X)", "E(X)", "E(X^2)", "E[(X-X.bar)^2]")
knitr::kable(output)
```

```{r echo=F}
cat("E(X):", sum(output[,3]), "\n")
cat("E(X^2):", sum(output[,4]), "\n")
cat("E[(X - X.bar)^2]:", sum(output[,5]), "\n")
```

---

__6.16__ A bag contains 2 white balls and 3 black balls. Each of four persons,
A, B, C, and D, in the order named, draws one ball and does not replace it. The
first to draw a white ball receives \$10. Determine the expectations of A, B, C,
and D.

__Solution__

Since only 3 black balls are present, one person must win on his or her first
attempt. Denote by A, B, C, and D the events "A wins," "B wins," "C wins," and
"D wins," respectively.

$$
\text{Pr\{A wins\}} = \text{Pr\{A\}} = \frac{2}{3 + 2} = \frac{2}{5}
$$

Thus A's expectation $= \frac{2}{5}(\$10) = \$4$.

$$
\text{Pr\{A loses and B wins\}} = \text{Pr}\{\bar{A} B\}
  = \text{Pr}\{\bar{A}\}\ \text{Pr}\{B|\bar{A}\}
  = \left(\frac{3}{5}\right) \left(\frac{2}{4}\right) = \frac{3}{10}
$$

Thus B's expectation = \$3.

$$
\text{Pr\{A and B lose and C wins\}} = \text{Pr}\{\bar{A} \bar{B} C\}
  = \text{Pr}\{\bar{A}\}\ \text{Pr}\{\bar{B}|\bar{A}\}\
    \text{Pr}\{C \bar{A} \bar{B}\}
  = \left(\frac{3}{5}\right) \left(\frac{2}{4}\right) \left(\frac{2}{3}\right)
  = \frac{1}{5}
$$

Thus C's expectation = \$2.

$$
\begin{aligned}
\text{Pr\{A, B, and C lose and D wins\}}
  & = \text{Pr}\{\bar{A}\bar{B}\bar{C}D\} \\
  & = \text{Pr}\{\bar{A}\}\ \text{Pr}\{\bar{B}|\bar{A}\}\ 
      \text{Pr}\{\bar{C}|\bar{A} \bar{B}\}\
      \text{Pr}\{D|\bar{A} \bar{B} \bar{C}\} \\ \\
  & = \left(\frac{3}{5}\right) \left(\frac{2}{4}\right)
      \left(\frac{1}{3}\right) \left(\frac{1}{1}\right) = \frac{1}{10}
\end{aligned}
$$

Thus D's expectation = \$1.

_Check_: \$4 + \$3 + \$2 + \$1 = \$10, and
$\frac{2}{5} + \frac{3}{10} + \frac{1}{5} + \frac{1}{10} = 1$.

---

__6.17__ In how many ways can 5 differently colored marbles be arranged in a
row?

__Solution__

We must arrange the 5 marbles in 5 positions: -- -- -- -- --. The first position
can be occupied by any one of 5 marbles (i.e., there are 5 ways of filling the
first position). When this has been done, there are 4 ways of filling the second
position. Then there are 3 way of filling the third position, 2 ways of filling
the fourth position, and finally only 1 way of filling the last position.
Therefore:

$$
\text{Number of arrangements of 5 marbles in a row }
  = 5 \cdot 4 \cdot 3 \cdot 2 \cdot 1 = 5! = 120
$$

In general,

$$
\text{Number of arrangements of n different objects in a row }
  = n(n-1)(n-2) \cdots 1 = n!
$$

This is also called th enumber of _permutations of n$ different objects taken
$n$ at a time and is denoted by $_{n}P_{n}$.

---

__6.18__ In how many ways can 10 people be seated on a bench if only 4 seats are
available?

__Solution__

The first seat can be filled in any one of 10 ways, and when this has been
done there are 9 ways of filling the second seat, 8 ways of filling the third
seat, and 7 ways of filling the fourth seat. Therefore:

$$
\text{Number of arrangements of 10 peiple taken 4 at a time }
  = 10 \cdot 9 \cdot 8 \cdot 7 = 5040
$$

In general

$$
\text{Number of arrangements of n different objects taken r at a time }
  = n(n-1) \cdots (n - r + 1) = n!
$$

This is also called the number of _permutations of n_ different objects taken
$r$ at a time and is denoted by $_{n}P_{r}$, $P(n, r)$ or $P_{n,r}$. Note
that when $r = n$, $_{n}P_{n} = n!$, as in Problem 6.17.

---

__6.19__ Evaluate (a) $_{8}P_{3}$, (b) $_{6}P_{4}$, (c) $_{15}P_{1}$, and
$_{3}P_{3}$, and (e) Parts (a) through (d).

__Solution__

(a) $_{8}P_{3} = 8 \cdot 7 \cdot 6 = 336$

(b) $_{6}P_{4} = 6 \cdot 5 \cdot 4 \cdot 3 = 360$

(c) $_{15}P_{1} = 15$

(d) $_{3}P_{3} = 3 \cdot 2 \cdot 1 = 6$

---

__6.20__ It is required to seat 5 men and 4 women in a row so that the women
occupy the even places. How many such arrangements are possible?

__Solution__

The men may be seated in $_{5}P_{5}$ ways and the women in $_{4}P_{4}$ ways;
each arrangement of the men may be associated with each arrangement of the
women. Hence the required number of arrangements is
$_{5}P_{5} \cdot _{4}P_{4} = 5!4! = (120)(24) = 2880$.

---

__6.21__ How many four-digit numbers can be formed with the 10 digits 0, 1, 2,
3, $\ldots$, 9, if (a) repetitions are allowed, (b) repetitions are not allowed,
and (c) the last digit must be zero and repetitions are not allowed?

__Solution__

(a) The first digit can be any ony one of 9 (since 0 is not allowed). The
second, third and fourth digits can be any one of 10. Then
$9 \cdot 10 \cdot 10 \cdot 10 = 9000$ numbers can be formed.

(b) The first digit can be any one of 9 (any one but 0). The second digit can be
any one of 9 (any but that used for the first digit). The third digit can be any
one of 8 (any but those used for the first two digits). The fourth digit can be
any one of 7 (any but those used for the first three digits). Thus
$9 \cdot 9 \cdot 8 \cdot 7 = 4536$ numbers can be formed.

(c) The first digit can be any one of 9 (any one but 0), the second in 8, and
third in 7. Thus $9 \cdot 8 \cdot 7 = 504$.

---

__6.22__ Four different mathematics books, 6 different physics books, and 2
different chemistry books are to be arranged on a shelf. How many different
arrangements are possible if (a) the books in each particular subject must all
stand together and (b) only the mathematics books must stand together?

__Solution__

(a) The mathematics books can be arranged among themselves in $_{4}P_{4}=4!$
ways, the physics books in $_{6}P_{6}=6!$ ways, the chemistry books in
$_{2}P_{2}=2!$ ways, and the three groups in $_{3}P_{3}=3!$ ways. Thus the
required number of arrangements = $4!6!2!3!=207,360$.

(b) Consider the 4 mathematics books as one big book. Then we have 9 books that
can be arranged in $_{9}P_{9}=9!$ ways. In all of these ways the mathematics
books can be arranged among themselves in $_{4}P_{4}=4!$ ways. Thus the required
number of arrangements $= 9!4! = 8,709,120$.

---

__6.23__ Five red marbles, 2 white marbles, and 3 blue marbles are arranged in
a row. If all the marbles of the same color are not distinguishable from each
other, how many different arrangements are possible?

__Solution__

Assume that there are P different arrangements. Multiplying P by the numbers
of ways of arranging (a) the 5 red marbles among themselves, (b) the 2 white
marbles among themselves, and (c) the 3 blue marbles among themselves (i.e.,
multiplying P by $5!2!3!$), we obtain the number of ways of arranging the 10
marbles if they are distinguisable (i.e., 10!). Thus

$$
(5!2!3!)P = 10!\ \ \ \text{ and }\ \ \ P = \frac{10!}{5!2!3!}
$$

In general, the number of different arrangements of n objects of which $n_1$ are
alike, $n_2$ are alike, $\ldots$, $n_k$ are alike is

$$
\frac{n!}{n_1! n_2! \cdots n_k!}
$$

where $n_1 + n_2 + \cdots + n_k = n$.

```{r}
factorial(10)/(factorial(5)*factorial(2)*factorial(3))
```

---

__6.24__ In how many ways can 7 people be seated at a round table if (a) they 
can sit anywhere and (b) 2 particular people must not sit next to each other?

__Solution__

(a) Let 1 of them be seated anywhere. Then the remaining 6 people can be seated
in $6! = 720$ ways, which is the total number of ways of arranging the 7 people
in a circle.

(b) Consider the 2 particular people as 1 person. Then there are 6 people
altogether and they can be arranged in $5!$ ways. But the 2 people considered as
1 can be arranged among themselves in $2!$ ways. Thus the number of ways of
arranging 6 people at a round table with 2 particular people sitting together
$= 5! 2! = 240$.

Then using part (a), the total number of ways in which 6 people can be seated at
a round table so that 2 particular people do not sit together $- 720 - 240 = 480$
ways.

















