---
title: "Elementary Probability Theory"
author: "Joseph Bennett"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    extra_dependencies:
      epsdice: null
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Fundamental rules of probability

## Definitions of probability: Classic versus Relative-frequency

__6.40.__ Determine the probability $p$, or an estimate of it, for each of the
following events:

(a) A king, ace, jack of clubs, or queen of diamonds appears in drawing a single
card from a well-shuffled ordinary deck of cards.
(b) The sum 8 appears in a single toss of a pair of fair dice.
(c) A non-defective bolt will be found if out of 600 bolts already examined, 12
were defective.
(d) A 7 or 11 comes up in a single toss of a pair of fair dice.
(e) At least one head appears in three tosses of a fair coin.

__Solution__

(a) The chance that a king (4 kings), ace (4 aces), a jack of clubs or a queen
of diamonds (10 cards in total) will be drawn is
$$
p = \frac{10}{52} = `r 10/52`\text{.}
$$

(b) The sum 8 can come up 5 out of 36 times:
\epsdice{6} \epsdice[black]{2}, 
\epsdice{5} \epsdice[black]{3},
\epsdice{4} \epsdice[black]{4},
\epsdice{3} \epsdice[black]{5},
\epsdice{2} \epsdice[black]{6},
$$
p = \frac{5}{36} = `r 5/36`\text{.}
$$

(c) From already examined bolts 12 were determined defective out of 600. The
probability that that a non defective bolt will be found is:
$$
p = 1 - 12/600 = `r 1 - 12/600`\text{.}
$$

(d) A 7 or 11 comes up with the toss of a pair of fair dice:

Seven

$$
\epsdice{6} \epsdice[black]{1}\ 
\epsdice{5} \epsdice[black]{2}\ 
\epsdice{4} \epsdice[black]{3}\ 
\epsdice{3} \epsdice[black]{4}\ 
\epsdice{2} \epsdice[black]{5}\ 
\epsdice{1} \epsdice[black]{6} = 6
$$
Eleven

$$
\epsdice{6} \epsdice[black]{5}\
\epsdice{5} \epsdice[black]{6} = 2
$$

$$
p = \frac{6 + 2}{36} = \frac{8}{36} = `r 8/36`\text{.}
$$

(e) The probability of at least one head appears in three tosses of a fair coin
is equal to 1 - the probability of no heads in three tosses.

$$
1 - \left(\frac{1}{2}\right) \left(\frac{1}{2}\right) \left(\frac{1}{2}\right)
 = `r 1 - (1/2)^3`
$$

---

## Probability nomenclature

__6.41.__ An experiment consists of drawing three cards in succession from a
well-shuffled ordinary deck of cards. Let $E_1$ be the event "king" on the first
draw, $E_2$ the event "king" on the second draw, and $E_3$ the event "king" on
the third draw. State in words the meaning of each of the following:

(a) Pr{$E_1 \bar{E_2}$}

(b) Pr{$E_1 + E_2$}

(c) $\bar{E_1} + \bar{E_2}$

(d) Pr{$E_3|E_1 \bar{E_2}$}

(e) $\bar{E_1} \bar{E_2} \bar{E_3}$

(f) Pr{$E_1 E_2 + \bar{E_2} E_3$}

__Solution__

(a) Pr{$E_1 \bar{E_2}$} - The probability of king on first draw and not king on
second draw.

(b) Pr{$E_1 + E_2$} - The probability of king on first draw or king on second
draw or both.

(c) $\bar{E_1} + \bar{E_2}$ - The event not king on first draw or not king on
second draw or both are not kings. Another way of stating this is at least one
of the first two draws is not a king.

(d) Pr{$E_3|E_1 \bar{E_2}$} - The probability of king on third draw given that
king on first draw and not king on second draw.

(e) $\bar{E_1} \bar{E_2} \bar{E_3}$ - The event not king on first draw and not
king on second draw and not king on third draw. Or no king is drawn in three
draws.

(f) Pr{$E_1 E_2 + \bar{E_2} E_3$} - The probability of king on first draw and
king on second draw or not king on second draw and king on third draw. Or the
probability that the first two draws are a king or the second draw is not a king
but the third draw is a king.

---

## Conditional probability, independent and dependent events, and mutually exclusive events

__6.42.__ A ball is drawn at random from a box containing 10 red, 30 white, 20
blue and 15 orange marbles. Find the probability that the ball drawn is (a)
orange or red, (b) not red or blue, (c) not blue, (d) white, and (e) red.

__Solution__

Summarizing the contents of the box:

- 10 Red marbles
- 30 White marbles
- 20 Blue marbles
- 15 Orange marbles
- 75 Total marbles

(a) $O + R = \frac{15 + 10}{75} = \frac{25}{75}$

(b) $\bar{R} + B = W + B + O  = \frac{30 + 20 + 15}{75} = \frac{65}{75}$

(c) $\bar{B} = \frac{10 + 30 + 15}{75} = \frac{55}{75}$

(d) $W = \frac{30}{75}$

(e) $R = \frac{10}{75}$

---

### Probability experiments with replacement

__6.43.__ Two marbles are drawn in succession from the box of Problem 6.42,
replacement being made after each drawing. Find the probability (a) both are
white, (b) the first is red and the second is white, (c) neither is orange, (d)
they are either red or white or both (red and white), (e) the second is not
blue, (f) the first is orange, (g) at least one is blue, (h) at most one is red,
(i) the first is white but the second is not, and (j) only one is red.

__Solution__

Summarizing the contents of the box:

- 10 Red marbles
- 30 White marbles
- 20 Blue marbles
- 15 Orange marbles
- 75 Total marbles

Since the selection of the marbles occur _with replacement_ we use the following
equation:

\begin{equation}
\text{Pr}\{E_1 E_2\} = \text{Pr}\{E_1\}\ \text{Pr}\{E_2\}
\end{equation}

(a) Both are white: 

$$
\text{Pr}\{W_1\}\ \text{Pr}\{W_2\} =
\left(\frac{30}{75}\right)^2 = `r (30/75)^2`
$$

(b) The first is red and the second is white:

$$
\text{Pr}\{R_1\}\ \text{Pr}\{W_2\} =
\left(\frac{10}{75}\right) \left(\frac{30}{75}\right) =
\left(\frac{300}{5625}\right) = `r (300/5625)`
$$

(c) Neither is orange:

$$
\text{Pr}\{\bar{O_1}\}\ \text{Pr}\{\bar{O_2}\}
= \left(\frac{60}{75}\right)^2 = `r (60/75)^2`
$$

(d) Either red or white or both:

$$
\text{Pr}\{R_1 + W_1\}\ \text{Pr}\{W_2 + R_2\}
= \left(\frac{10 + 30}{75}\right)^2 = `r (40/75)^2`
$$

(e) The second is not blue:

$$
\text{Pr}\{\bar{B_2}\} = \text{Pr}\{R_2 + W_2 + O_2\} = \frac{10 + 30 + 15}{75} = `r 55/75`
$$

(f) The first is orange:

$$
\text{Pr}\{O_1\} = \frac{15}{75} = `r 15/75`
$$

(g) At least one is blue is the same as 1 - Pr{no blue}:

$$
1 - \left(\text{Pr}\{\bar{B_1}\}\ \text{Pr}\{\bar{B_2}\}\right)
= 1 - \left(\frac{55}{75}\right)^2 = `r 1 - (55/75)^2`
$$

(h) At most one red is the same as 1 - Pr{all red}:

$$
1 - \left(\text{Pr}\{R_1\}\ \text{Pr}\{R_2\}\right)
= 1 - \left(\frac{10}{75}\right)^2 = `r 1 - (10/75)^2`
$$

(i) The first is white but the second is not white:

$$
\text{Pr}\{W_1\}\ \text{Pr}\{\bar{W_2}\} = 
\left(\frac{30}{75}\right) \left(\frac{45}{75}\right) = `r (30/75)*(45/75)`
$$

(j) Only one is red:

$$
\text{Pr}\{R_1\} + \text{Pr}\{R_2\}
    - \left(\text{Pr}\{R_1\}\ \text{Pr}\{R_2\}\right)
= 2 * \left(\frac{10}{75}\right) - \left(\frac{10}{75}\right)^2
= `r 2*(10/75) - (10/75)^2`
$$

---

### Probability experiments without replacement

__6.44.__ Work Problem 6.43 if there is no replacement after each drawing.

__Solution__

Summarizing the contents of the box:

- 10 Red marbles
- 30 White marbles
- 20 Blue marbles
- 15 Orange marbles
- 75 Total marbles

Since the marbles are selected from the box _without replacement_, the events
are not independent. The second event depends on the first. We take this into
account by using the following equation:

\begin{equation}
\text{Pr}\{E_1 E_2\} = \text{Pr}\{E_1\}\ \text{Pr}\{E_2|E_1\}
\end{equation}

The component $\text{Pr}\{E_2|E_1\}$ is executed by removing the marble from
the equation.

---

(a) Both are white:

$$
\text{Pr}\{W_1\}\ \text{Pr}\{W_1|W_2\}
= \left(\frac{30}{75}\right) \left(\frac{29}{74}\right)
= `r (30/75)*(29/74)`
$$

(b) The first is red and the second is white

$$
\text{Pr}\{R_1\}\ \text{Pr}\{W_2|R_1\}
= \left(\frac{10}{75}\right) \left(\frac{30}{74}\right)
= `r (10/75)*(30/74)`
$$

(c) Neither is orange

$$
\text{Pr}\{\bar{O_1}\}\ \text{Pr}\{\bar{O_2}|\bar{O_1}\}
= \left(\frac{60}{75}\right) \left(\frac{59}{74}\right)
= `r (60/75)*(59/74)`
$$

(d) Both are red or white or both

$$
\text{Pr}\{R_1 + W_1\}\ \text{Pr}\{(R_2 + W_2)|(R_1 + W_1)\}
= \left(\frac{10 + 30}{75}\right) \left(\frac{10 + 30 - 1}{74}\right)
= `r (40/75)*(39/74)`
$$

(e) The second is not blue

It doesn't matter what the first draw is therefore we are only concerned with
the second draw.

$$
\text{Pr}\{\bar{B_2}\} = \left(\frac{75 - 20}{75}\right) = `r (75-20)/75`
$$

(f) The first is orange

It doesn't matter what the second draw marble is, we are only concerned with
the first draw.

$$
\text{Pr}\{O_1\} = \frac{15}{75} = `r 15/75`
$$

(g) At least one is blue is the same as 1 - Pr{no blue}.

$$
1 - \left(\text{Pr}\{\bar{B_1}\}\ \text{Pr}\{\bar{B_2}|\bar{B_1}\}\right)
= 1 - \left(\frac{55}{75}\right) \left(\frac{54}{74}\right)
= `r 1 - (55/75)*(54/74)`
$$

(h) At most one is red is the same as 1 - Pr{all red}.

$$
1 - \left(\text{Pr}\{R_1\}\ \text{Pr}\{R_2|R_1\}\right)
= \left(\frac{10}{75}\right) \left(\frac{9}{74}\right)
= `r 1- (10/75)*(9/74)`
$$

(i) The first is white but the second is not.

$$
\text{Pr}\{W_1\}\ \text{Pr}\{\bar{W_2}|W_1\}
= \left(\frac{30}{75}\right) \left(\frac{45}{74}\right)
= `r (30/75)*(45/74)`
$$

(j) Only one is red.

$$
\text{Pr}\{R_1\}\ \text{Pr}\{\bar{R_2}|R_1\} + \text{Pr}\{\bar{R_1}\}\ \text{Pr}\{\bar{R_1}|R_2\}
= \left(\frac{10}{75}\right) \left(\frac{64}{74}\right)
+ \left(\frac{65}{75}\right) \left(\frac{10}{74}\right)
= `r (10/75)*(64/74)+(65/75)*(10/74)`
$$

---

## Probability experiments: probability space

__6.45.__ Find the probability of scoring a total of 7 points (a) once, (b) at
least once, and (c) twice in two tosses of a pair of fair dice.

__Solution__

To solve these probabilities, we need to understand the ways a total of
_7 points_ can appear on a single toss of a pair of fair dice. 

The sum of 7 can be achieved with the following pairs:
- (1, 6), (6, 1)
- (2, 5), (5, 2)
- (3, 4), (4, 3)

$$
\text{Pr}\{\text{7 in one toss}\} = \frac{6}{36} = \frac{1}{6}.
$$

(a) The probability of scoring exactly 7 once:

$$
\text{Pr}\{E_1 \bar{E_2}\} + \text{Pr}\{\bar{E_1} E_2\}
= 2 \times \left(\frac{1}{6}\right) \left(\frac{5}{6}\right)
= `r 2 *(5/36)`
$$

(b) The probability of scoring 7 at least once in two tosses is the same
as 1 - Pr{no 7}.

$$
1 - \text{Pr}\{\text{no 7}\} = 1 - \left(\frac{5}{6}\right)^2
= `r 1-(5/6)^2`
$$

(c) Probability of scoring 7 twice

$$
\text{Pr}\{E_1\}\ \text{pr}\{E_2\} = \left(\frac{1}{6}\right)^2
= `r (1/6)^2`
$$

---

__6.46.__ Two cards are drawn successively from an ordinary deck of 52
well-shuffled cards. Find the probability that (a) the first card is not a 10 of
clubs or an ace, (b) the first card is an ace but the second is not, (c) at
least one card is a diamond, (d) the cards are not of the same suit, (e) not
more than one card is a picture card (jack, queen, king), (f) the second card is
not a picture card, (g) the second card is not a picture card given that the
first was a picture card, and (h) the cards are picture cards or spades or both.

__Solution__

Let's solve each part step-by-step. We're working with a standard deck of **52 cards**. When cards are drawn **successively without replacement**, the total number of available cards decreases after the first draw, affecting the probability of the second draw.

(a) The first card is not a 10 of clubs or an ace:

There are five card total (1 10 of clubs and 4 aces).

$$
\text{Pr}\{\bar{E_1} + \bar{F_1}\} = `r 1 - 5/52`
$$

(b) The first card is an ace but the second is not

$$
\text{Pr}\{E_1 \bar{E_2}\} = \text{Pr}\{E_1\}\ \text{Pr}\{\bar{E_2}|E_1\}
\left(\frac{4}{52}\right) \left(1 - \frac{3}{51}\right) = `r 4/52 * (1 - 3/51)`
$$

(c) At least one card is a diamond is the same as 1 - Pr{both are not diamond}

$$
1 - \text{Pr}\{\bar{E_1}\}\ \text{Pr}\{\bar{E_2}|\bar{E_1}\} = 1 
- \left(\frac{52-13}{52}\right) \left(\frac{51-13}{51}\right)
= `r 1 - (52-13)/52*(51-13)/51`
$$

(d) The cards are not the same suit is the same as 1 -
Pr{both are the same suit}

The probability of drawing a card from a suit is 1 therefore
$\text{Pr}\{E_1\} = 1$

$$
1 - \text{Pr}\{E_1\}\ \text{Pr}\{E_2|E_1\}
= 1 - \left(\frac{12}{51}\right)
= `r 1- (12/51)`
$$


(e) The probability that not more than one card is a picture card (jack, queen,
king).

Two cases: (1) exactly one face card in two draws, or no face cards in two
draws.

Exactly one face card in two draws:

$$
\begin{aligned}
\text{Pr}\{E_1 \bar{E_2}\} + \text{Pr}\{\bar{E_1} E_2\}
& = \text{Pr}\{E_1\}\ \text{Pr}\{\bar{E_2}\} +
  \text{Pr}\{\bar{E_1}\}\ \text{Pr}\{E_2\} \\ \\
& = \left(\frac{12}{52}\right) \left(\frac{52-12}{51}\right) +
  \left(\frac{52-12}{52}\right) \left(\frac{12}{51}\right) \\ \\
& = `r (12/52)*(40/51) + (40/52)*(12/51)`
\end{aligned}
$$

Neither card is a face card in two draws:

$$
\text{Pr}\{\bar{E_1} \bar{E_2}\} =
\text{Pr}\{\bar{E_1}\}\ \text{Pr}\{\bar{E_2}\} = 
\left(\frac{40}{52}\right) \left(\frac{39}{51}\right) =
`r (40/52) * (39/51)`
$$

Therefore the probability is
$$
`r (p6.46e1 <- (12/52)*(40/51)+(40/52)*(12/51))` +
`r (p6.46e2 <- (40/52)*(39/51))`
= `r p6.46e1 + p6.46e2`
$$

(f) The second card is not a picture card. It doesn't matter what the first
card is

$$
\text{Pr}\{\bar{E_2}\} = \frac{40}{52} = `r 40/52`
$$

(g) The second card is not a picture card, given that the first was a picture
card:

$$
\text{Pr}\{\bar{E_2}|E_1\}
= \left(\frac{40}{51}\right)
= `r (40/51)`
$$
(h) The cards are picture cards or spades or both


$$
\text{Pr}\{E_1\}\ \text{Pr}\{E_2|E_1\}
= \left(\frac{13 + 12 - 2 = 22}{52}\right) \left(\frac{21}{51}\right)
= `r 22/52 * 21/51`
$$

---

__6.47.__ A box contains 9 tickets numbered from 1 to 9 inclusive. If 3 tickets are drawn from the box one at a time, find the probability that they are alternately either (1) odd, even, odd or (2) even, odd, even.

__Solution__

From 1 to 9 there are 5 odd numbers and 4 even numbers.

(1) For odd-even-odd

$$
\text{Pr}\{O_1 E_2 O_3\} = \text{Pr}\{O_1\}\ \text{Pr}\{E_2|O_1\}\ 
\text{Pr}\{O_3|O_1 E_2\} =
\left(\frac{5}{9}\right) \left(\frac{4}{8}\right) \left(\frac{4}{7}\right)
= `r (5/9)*(4/8)*(4/7)`
$$

(2) For even-odd-even

$$
\text{Pr}\{E_1 O_2 E_3\} = \text{Pr}\{E_1\}\ \text{Pr}\{O_2|E_1\}\ 
\text{Pr}\{E_3|E_1 O_2\} = 
\left(\frac{4}{9}\right) \left(\frac{5}{8}\right) \left(\frac{3}{7}\right) =
`r (4/9) * (5/8) * (3/7)`
$$

---

__6.48.__ The odds in favor of A winning a game of chess against B are 3:2. If
three games are to be played, what are the odds (a) in favor of A's winning at
least two games out of the three and (b) against A losing the first two games to
B?

__Solution__

The probability of A winning against B is Pr{A wins} = $\frac{3}{3 + 2} = 0.6$.

The probability of B winning against A is Pr{B wins} = Pr{not A Wins} 
= $\frac{2}{3 + 2} = 1 - 0.6 = 0.4$

(a) The odds of A winning at least two games out of three is the same as the
chance of winning 2 games and the chance of winning all three games:

$$
\begin{aligned}
\text{Pr}\{\text{win 2 out of 3 games}\} & =
3 \times \text{Pr}\{A_1\}\ \text{Pr}\{A_2\}\ \text{Pr}\{\bar{A_3}\}
= 3 \times \left(\frac{3}{5}\right) \left(\frac{3}{5}\right)
   \left(\frac{2}{5}\right) = \frac{54}{125} \\ \\
\text{Pr}\{\text{win all 3 games}\} & = \text{Pr}\{A_1\}\ \text{Pr}\{A_2\}\ 
   \text{Pr}\{A_3\} = \left(\frac{3}{5}\right)^2 = \frac{27}{125} \\ \\
\text{Pr}\{\text{not winning at least 2 games}\} & =
    1 - \frac{81}{125} = \frac{125 - 81}{125} = \frac{44}{125}
\end{aligned}
$$

Odds in favor of winning at least two out of three games is $81:44$.

(b) Odds against A losing the first two games to B.

$$
\begin{aligned}
\text{Pr}\{\text{losing the first two games}\} & =
    \left(\frac{2}{5}\right) \left(\frac{2}{5}\right) = \frac{4}{25} \\ \\
\text{Pr}\{\text{not losing the first two games}\} & =
1 - \text{Pr}\{\text{losing the first two games}\} = \frac{25 - 4}{25} =
\frac{21}{25}
\end{aligned}
$$

The odds against A losing the first two games is $21:4$.

---

__6.49.__ A purse contains 2 silver coins and 4 copper coins, and a second purse contains 4 silver coins and 3 copper coins. If a coin is selected at random from one of the two purses, what is the probability that it is a silver coin?

__Solution__

Purse 1: 2 silver; total 6 coins

Purse 2: 4 silver; total 7 coins

$$
\begin{aligned}
\text{Pr}\{\text{selecting a silver coin from purse 1}\} = \frac{2}{6}\\ \\
\text{Pr}\{\text{selecting a silver coin from purse 2}\} = \frac{4}{7} \\ \\
\text{Pr}\{\text{silver}\} = \frac{1}{2} \times
    \left(\frac{2}{6} + \frac{4}{7}\right) = \frac{19}{42}
\end{aligned}
$$

---

__6.50.__ The probability that a man will be alive in 25 years is $\frac{3}{5}$,
and the probability that his wife will be alive in 25 years is $\frac{2}{3}$.
Find the probability that (a) both will be alive, (b) only the man will be alive,
(c) only the wife will be alive, and (d) at least one will be alive.

__Solution__

(a)

$$
\text{Pr}\{\text{both alive}\} = \text{Pr}\{\text{man alive}\}\ 
    \text{Pr}\{\text{woman alive}\}  = \frac{3}{5} \times \frac{2}{3} =
    \frac{6}{15}
$$

(b)

$$
\text{Pr}\{\text{only the man}\} = \frac{3}{5} \times (1 - \frac{2}{3})
  = \frac{3}{15}
$$

(c)

$$
\text{Pr}\{\text{only the wife}\} = (1 - \frac{3}{5}) \times \frac{2}{3} =
  \frac{4}{15}
$$

(d)

$$
\text{Pr}\{\text{at least one alive}\} = 1 - \left(\frac{2}{5} \times \frac{1}{3}\right)
  = 1 - \frac{2}{15} = \frac{13}{15}
$$

---

__6.51.__ Out of 800 families with 4 children each, what percentage would be expected to have (a) 2 boys and 2 girls, (b) at least 1 boy, (c) no girls, and (d) at most 2 girls? Assume equal probabilities for boys and girls.

__Solution__

To solve the problem, we'll use the binomial probability formula. The situation involves 4 children per family, and we will denote:

- The probability of having a boy (\(B\)) or a girl (\(G\)) is equal:  
  \[
  P(B) = P(G) = \frac{1}{2}
  \]

Let \( n = 4 \) (number of children) and \( p = \frac{1}{2} \) (probability of having a boy).

The binomial probability formula is given by:
\[
P(X = k) = \binom{n}{k} p^k (1-p)^{n-k}
\]
where:
- \( n \) = total number of trials (children)
- \( k \) = number of successes (boys or girls)
- \( p \) = probability of success (having a boy or girl)

---

### (a) Probability of 2 boys and 2 girls

For this part, we want to find the probability of having exactly 2 boys (\(k = 2\)):
\[
P(X = 2) = \binom{4}{2} \left(\frac{1}{2}\right)^2 \left(\frac{1}{2}\right)^{4-2}
\]
Calculating it:
\[
P(X = 2) = \binom{4}{2} \left(\frac{1}{2}\right)^4 = 6 \cdot \frac{1}{16} = \frac{6}{16} = \frac{3}{8}
\]

**Expected percentage in 800 families**:
\[
\text{Expected} = P(X = 2) \times 800 = \frac{3}{8} \times 800 = 300
\]

**Percentage**:
\[
\text{Percentage} = \frac{300}{800} \times 100 = 37.5\%
\]

---

### (b) Probability of at least 1 boy

The probability of at least 1 boy can be found by calculating the complement (the probability of having no boys):
\[
P(\text{at least 1 boy}) = 1 - P(\text{no boys}) = 1 - P(X = 0)
\]
For \( k = 0 \):
\[
P(X = 0) = \binom{4}{0} \left(\frac{1}{2}\right)^0 \left(\frac{1}{2}\right)^{4} = 1 \cdot 1 \cdot \frac{1}{16} = \frac{1}{16}
\]
Thus:
\[
P(\text{at least 1 boy}) = 1 - \frac{1}{16} = \frac{15}{16}
\]

**Expected percentage in 800 families**:
\[
\text{Expected} = P(\text{at least 1 boy}) \times 800 = \frac{15}{16} \times 800 = 750
\]

**Percentage**:
\[
\text{Percentage} = \frac{750}{800} \times 100 = 93.75\%
\]

---

### (c) Probability of no girls

The probability of having no girls is the same as having all boys:
\[
P(\text{no girls}) = P(X = 4) = \binom{4}{4} \left(\frac{1}{2}\right)^4 = 1 \cdot \frac{1}{16} = \frac{1}{16}
\]

**Expected percentage in 800 families**:
\[
\text{Expected} = P(\text{no girls}) \times 800 = \frac{1}{16} \times 800 = 50
\]

**Percentage**:
\[
\text{Percentage} = \frac{50}{800} \times 100 = 6.25\%
\]

---

### (d) Probability of at most 2 girls

The probability of having at most 2 girls is the sum of probabilities of having 0, 1, or 2 girls:
\[
P(X \leq 2) = P(X = 0) + P(X = 1) + P(X = 2)
\]

- \( P(X = 0) = \frac{1}{16} \) (calculated before)
- \( P(X = 1) \):
\[
P(X = 1) = \binom{4}{1} \left(\frac{1}{2}\right)^1 \left(\frac{1}{2}\right)^{4-1} = 4 \cdot \frac{1}{2} \cdot \frac{1}{8} = \frac{4}{16} = \frac{1}{4}
\]
- \( P(X = 2) = \frac{3}{8} \) (calculated before)

Now, combining these:
\[
P(X \leq 2) = \frac{1}{16} + \frac{1}{4} + \frac{3}{8}
\]

Convert \( \frac{1}{4} \) and \( \frac{3}{8} \) to sixteenths:
- \( \frac{1}{4} = \frac{4}{16} \)
- \( \frac{3}{8} = \frac{6}{16} \)

So:
\[
P(X \leq 2) = \frac{1}{16} + \frac{4}{16} + \frac{6}{16} = \frac{11}{16}
\]

**Expected percentage in 800 families**:
\[
\text{Expected} = P(X \leq 2) \times 800 = \frac{11}{16} \times 800 = 550
\]

**Percentage**:
\[
\text{Percentage} = \frac{550}{800} \times 100 = 68.75\%
\]

---

### Summary of Results

1. **(a)** Expected percentage of families with 2 boys and 2 girls: **37.5%**
2. **(b)** Expected percentage of families with at least 1 boy: **93.75%**
3. **(c)** Expected percentage of families with no girls: **6.25%**
4. **(d)** Expected percentage of families with at most 2 girls: **68.75%**

---

# Probability distributions

__6.52__ If $X$ is the random variable showing the number of boys in families
with  4 children (see Problem 6.51), (a) construct a table showing the
probability distribution of X and (b) represent the distribution in part (a)
graphically.

__Solution__

Using `R` `dbinom` function we can construct the table as such:

```{r p6_52a, echo=F, message=F, fig.cap="Table"}
require(ggplot2)
Boys <- c(0,1,2,3,4)
p.X <- dbinom(Boys,4,.5)
df <- data.frame(Boys,p.X)
colnames(df) <- c("X", "p(X)")
knitr::kable(df, caption="Table")
```


```{r p6_52b, echo=F, message=F, fig.cap="Plot"}
ggplot(df, aes(x=Boys, y=p.X)) + geom_col()
```

---

__6.53__ A continuous random variable $X$ that can assume values only between
$X=2$ and 8 inclusive has a density function given by $a(X+3)$, where $a$ is a
constant, (a) Calculate $a$. Find (b) Pr{3 < X < 5}, (c) Pr{X $\ge$ 4}, and
(d) Pr{$|X-5| < 0.5$}.

__Solution__

(a) Note that the area under this curve must equal 1.

$$
\begin{aligned}
\int^8_2{a(X + 3)}dx = a \left[\frac{1}{2}X^2 + 3X \right]^8_2 = 1 \\ \\
  a\left(\left(\frac{1}{2}8^2 + 3(8)\right)
  - \left(\frac{1}{2}2^2 + 3(2)\right)\right) = 1 \\ \\
48a = 1 \\ \\
a = \frac{1}{48}
\end{aligned}
$$

```{r p6.53}
require(ggplot2)
X <- 2:8
y <- (1/48)*(X + 3)
df <- data.frame(X,y)
ggplot(df, aes(x=X, y=y)) +
  geom_line() +
  geom_area( colour = "black", fill = "lightblue")
```

(b) To find the probability between 3 and 5: Pr{3 < X < 5}

$$
\begin{aligned}
\frac{1}{48}\int^5_3{(X + 3)}dx
  & = \frac{1}{48} \left[\frac{1}{2}X^2 + 3X \right]^5_3 \\ \\
  & = \frac{1}{48} \left(\left(\frac{1}{2}5^2 + 3(5)\right)
    - \left(\frac{1}{2}3^2 + 3(3)\right)\right) \\ \\
  & = `r (1/48)*((.5*5^2 + 3*5) - (.5*3^2 + 3*3))`
\end{aligned}
$$

(c) Pr{X $\ge$ 4}

$$
\frac{1}{48}\int^{\infty}_4{(X+3)}dx = \frac{1}{48}\int^8_4{(X + 3)}dx = 
`r (1/48)*((.5*8^2 + 3*8) - (.5*4^2 + 3*4))`
$$

(d) Pr{$|X-5| < 0.5$}


$$
\begin{aligned}
|X - 5| < 0.5 \\ \\
-0.5 < X - 5 < 0.5 
\end{aligned}
$$

Solving for $X$

$$
4.5 < X < 5.5
$$

$$
\frac{1}{48} \int^{5.5}_{4.5}{X-3}dx
  = \frac{1}{48}\left[{\frac{1}{2}X^2 + 3X} \right]^{5.5}_{4.5}
  = `r 1/48*((.5*5.5^2 + 3 * 5.5) - (.5*4.5^2 + 3 * 4.5))`
$$

---

__6.54__ Three marbles are drawn without replacement from an urn containing 4
red and 6 white marbles. If $X$ is a random variable that denotes the total
number of red marbles drawn, (a) construct a table showing the probability
distribution of X and (b) graph the distribution.

__Solution__

The urn contains 4 red and 6 white marbles for a total of 10. The probability
of drawing a red marble is $0.4$

```{r p6.54}
require(ggplot2)
X <- 0:3
p.X <- dhyper(X,4,6,3)
(p6.54.df <- data.frame(X, p.X))
ggplot(p6.54.df, aes(x=X, y=p.X)) + geom_col()
```

---

__6.55__

(a) Three dice are rolled and $X=$ the sum on the three upturn faces. Give the
probability distribution for X. (b) Find Pr{7 $\le$ X $\le$ 11}.

__Solution__

(a) To derive the probability distribution we need to define the outcomes. The
set of outcomes range from 3 to 18. Below is a table of the number of
combinations for each result.

```{r p6.55a, fig.cap="Probability distribution", message=F}
require(data.table)
require(ggplot2)

# Number of simulations
outcomes <- 6^3

# All possible sum of rolling 3 dice
sums <- rowSums(expand.grid(1:6, 1:6, 1:6))

# Distribution table
p6.55.dt <- as.data.table(table(sums) / outcomes)
colnames(p6.55.dt) <- c("Sum", "Probability")
p6.55.dt$Sum <- as.numeric(p6.55.dt$Sum)

knitr::kable(p6.55.dt, caption="Probability distribution")
ggplot(p6.55.dt, aes(x=Sum, y=Probability)) + geom_col()
```

(b) Pr{7 $\le$ X $\le$ 11}

```{r p6.55b}
sum(p6.55.dt[Sum >= 7 & Sum <= 11]$Probability)
```

---

# Mathematical expectation

__6.56__ What is a fair price to pay to enter a game in which one can win \$25
with probability 0.2 and \$10 with probability 0.4?

$$
\text{Expected value is } (0.2)(\$25) + (0.4)(\$10) = \$`r .2*25 + .4*10`
$$

---

__6.57__ If it rains, an umbrella salesman can earn \$30 per day. If it is fair,
he can lose \$6 per day. What is his expectation if the probability of rain is
0.3?

$$
\text{Expected value is } (0.3)(\$30) - (0.7)(\$6) = \$`r .3*30 - .7*6`
$$

---

__6.58__ $A$ and $B$ play a game in which they toss a fair coin 3 times. The
one obtaining heads first wins the game. If $A$ tosses the coin first and if
the total value of the stakes is \$20, how much should be contributed by each in
order that the game be considered fair?

The probability that A wins on the first round:

$$
\text{Pr\{A wins on first\}} = \text{Pr}\{A_1\} = \frac{1}{2}
$$

The probability that B wins on the second round:

$$
\text{Pr\{B wins second\}} = \text{Pr}\{\bar{A_1}\}\ \text{Pr}\{B_2|\bar{A_1}\}
  = \left(\frac{1}{2}\right) \left(\frac{1}{2}\right) = \frac{1}{4}
$$

The probability that A wins on the third round:

$$
\text{Pr\{A wins on third\}} = \text{Pr}\{A_3|\bar{B_2} \bar{A_1}\}
  = \text{Pr}\{A_3\}\ \text{Pr}\{\bar{B_2}|\bar{A_1}\}
  = \left(\frac{1}{2}\right) \left(\frac{1}{4}\right) = \frac{1}{8}
$$

To make the game fair, the contributions should be proportioned as such: A
contributes \$`r (1/2+1/8)*20` and B contributes \$`r (1/4)*20`.

---

__6.59__ Find (a) $E(X)$, (b) $E(X^2)$, (c) $E[(X - \bar{X})^2]$, and 
(d) $E(X^3)$ for the probability distribution of Table 6.4.

```{r p6.59.table, echo=F, warning=F, message=F}
require(dplyr)
require(knitrExtra)
X <- c(-10, -20, 30)
p.X <- c(1/5, 3/10, 1/2)
df <- data.frame(X,p.X)
df
```

__Solution__

```{r p6.59sol, echo=F, warning=F, message=F}
sum(E.X <- df$X*df$p.X)
sum(E.X.sq <- df$X^2*df$p.X)
sum(V.X <- df$p.X*(df$X - mean(df$X))^2)
sum(E.X.cu <- df$X^3*df$p.X)
```

---

# Relation between population, sample mean, and variance

__6.60__ Referring to Problem 6.54, find the (a) mean, (b) variance, and (c)
standard deviation of the distribution of X, and interpret your results.

__Solution__

(a) Assuming that the
_expected value of the sample mean m is the population mean $\mu$_

```{r p6.60a}
(X <- p6.54.df$X)
(p.X <- p6.54.df$p.X)
(E.X <- sum(p.X*X))
(X.var <- sum(p.X*(X - E.X)^2))
(X.sd <- sqrt(X.var))
```

---

__6.61__ A random variable assumes the value 1 with probability $p$, and 0 with
probability $q=1-p$. Prove that

(a) $E(X) = p$ and (b) $E[(X-\bar{X})^2]=pq$

__Solution__

```{r p6.61}
X <- c(0, 1)
p.X <- c('q', 'p')
```

$$
E(X) = \sum{pX} = (1-p)(0) + p(1) = p \\ \\
E[(X - \bar{X})^2] = E(X^2) - [E(X)]^2 = p - p^2 = p(1-p) = pq
$$

---

__6.62__ Prove that (a) $E(2X + 3) = 2E(X) + 3$ and (b)
$E[(X - \bar{X})^2] = E(X^2) - [E(X)]^2$

(a)

$$
E(2X + 3) = E(2X) + E(3) = 2E(X) + 3
$$

(b)

$$
\begin{aligned}
E[(X - \bar{X})^2] & = E[(X - E(X))^2] \\
  & = E[X^2 - 2XE(X) + E(X)^2] \\
  & = E(X^2) - 2[E(X)]^2 + [E(X)]^2 \\
  & = E(X^2) - [E(X)]^2
\end{aligned}
$$

---

__6.63__ In Problem 6.55, find the expected value of X.

__Solution__

```{r p6.63}
dt <- p6.55.dt
(E.X <- sum(dt$Probability*dt$Sum))
```

---

# Permutations

```{r combinat, message=F, echo=F, warning=F}
perm <- function(n, k) { choose(n, k) * factorial(k) }
```

__6.64__ Evaluate (a) $_{4}P_{2}$, (b) $_{7}P_{5}$, and (c) $_{10}P_{3}$.

```{r}
perm(4, 2)
perm(7, 5)
perm(10, 3)
```

---

__6.65__ For what value of $n$ is $_{n+1}P_{3} = _{n}P_{4}$?

__Solution__

$$
\begin{aligned}
\frac{(n+1)!}{((n+1) - 3)!} & = \frac{n!}{(n - 4)!} \\ \\
\frac{(n+1) \cdot n!}{(n-2)!} & = \frac{n!}{(n-4)!} \\ \\
\frac{(n+1)(n-4)!}{(n-2)!} & = 1 \\ \\
\frac{(n+1)(n-4)!}{(n-2)(n-3)(n-4)!} & = 1 \\ \\
\frac{(n+1)}{(n-2)(n-3)} & = 1 \\ \\
n + 1 & = n^2 - 5n + 6 \\
0 & = n^2 - 6n + 5 \\ \\
n = 5\ \ \ & \text{ or }\ \ \ n = 1 \\
n = 5
\end{aligned}
$$

---

__6.66__ In how many ways can 5 people be seated on a sofa if there are only 3
seats available?

__Solution__

The answer is to find $_{5}P_{3}$:

```{r p6.66}
perm(5, 3)
```

---

__6.67__ In how many ways can 7 books be arranged on a shelf if (a) any
arrangement is possible, (b) 3 particular books must always stand together,
and (c) 2 particular books must occupy the ends?

__Solution__

(a) For any arrangement:

```{r p6.66a}
factorial(7)
```

(b) If 3 books must be together:

```{r p6.66b}
factorial(3) * factorial(5)
```

(c) If 2 particular books must occupy the ends:

```{r p6.66c}
2 * factorial(5)
```

---

__6.68__ How many numbers consisting of five different digits each can be made
from the digits 1, 2, 3, $\ldots$, 9 if (a) the numbers must be odd and (b)
the first two digits of each number are even?

__Solution__

There are 5 odd numbers and 4 even numbers between 1--9 inclusive.

(a) If the numbers must be odd then the last digit of the number must be odd.
After the last number has been selected, there are 8 numbers and four slots
remaining.

```{r p6.68a}
5 * perm(8,4)
```

(b) If the first two digits of each number are even, then there are 12
permutations of those two numbers using even numbers, then there are 7 numbers
and 3 slots remaining.

```{r p6.68b}
perm(4, 2) * perm(7, 3)
```

---

__6.69__ Solve Problem 6.68 if repetitions of the digits are allowed.

(a) The total number of ways to arrange the first four digits, with the last
digit being odd (5 numbers)

```{r p6.69a}
5 * 9^4
```


(b) If the first two numbers are even:

```{r p6.69b}
4^2 * 9^3
```

---

__6.70__ How many different three-digit numbers can be made with three 4's,
four 2's, and two 3's?

__Solution__

We have three cases for this:

1. All three numbers are the same.
2. Two of the number are the same and the last is different.
3. All three numbers are different.

Case 1: There are only 2 three digit numbers that can be made where all three
numerals are the same: 444 and 222.

Case 2: If 2 numbers are the same then we have:

- six numbers with two 4's: 442, 424, 244, 443, 434, 344.
- six numbers with two 2's: 223, 232, 322, 224, 242, 422.
- six numbers with two 3's: 334, 343, 433, 332, 323, 233.

Therefore there are 18 numbers in case 2.

Case 3: All three numbers there are `r 3*2*1` permutations of 432.

There are $2 + 18 + 6 = 26$ different three digit numbers.

---

__6.71__ In how many ways can 3 men and 3 women be seated at a round table if 
(a) no restrictions are imposed (b) 2 particular women must not sit together,
and (c) each woman is to be between 2 men?

__Solution__

(a) The first one will be seated then `r 5*4*3*2*1` permutations.

(b) The two will be seated then `r 4*3*2*1` permutations.

(c) A woman is seated and 

Therefore there are `r 2+3*2*1` different three-digit numbers







